Practical consistency management for geographically distributed MMOG servers

### 2011-05-26 trunk/textos/optpaxos/draft.tex
quasi-genuine global fifo total order multicast pronto

@@

As messages are delivered according to their generation time, the delivery order is FIFO. Besides, as every two messages which are delivered in different groups should be delivered in the same order in these groups, we need total order. However, FIFO total order is not enough to guarantee that a message $m_1$ is delivered before a message $m_2$, which came from a different source than $m_1$, if the time $t_1$, at which $m_1$ was generated at the process $p_1$ which sent it, is such that $t_1 < t_2$, where $t_2$ is the time at which $m_2$ was generated at $p_2$, according to $p_2$'s clock, where $p_2$ is the source of $m_2$.

Although this property may be hard to observe, it is used by the optimistic delivery protocol, so that messages can be delivered in order in one communication step, in the best case. If, when each process sends a message $m$, it stamps the message with the current value of its wallclock, which will be $st(m)$, we define:

\textbf{Global FIFO order}: if processes $p_i$ and $p_j$ send respectively $m_i$ and $m_j$ to the same destination $p_d$, then if $st(m_i) < st(m_j)$ then $m_j$ is not delivered after $m_i$ in $p_d$.

@@

The basic idea of the optimistic ordering is the following: assuming that the servers have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enough, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a server $S$ (either coordinator or replica) receives a message $m$ from a client, it immediately applies a timestamp $ts$ for it, which consists simply of the current value of the server's wallclock, $now$. Therefore, $m.ts=now$. A wait window of length $w(S)$ is considered, where $w(S)$ is defined as the highest estimated communication delay between the server $S$ and any of the other servers in its group or in its group's neighbors.

After applying the timestamp to $m$, the server $S$ immediately forwards it to all the other servers involved, including servers in other groups. A server is involved with a message $m$ when it is one of its destinations, in $m.dst$. Then, $S$ puts $m$ on a list (\mbox{\textit{optPending}}), where it stays until $now>m.ts+w(S)$, which means that $m$ has been in that list for a time longer than the defined wait window. In the meantime, other messages, sent from other clients and forwarded by other servers, may have been received and also inserted in that list, always sorting by the timestamp of their timestamps. If $w(S)$ has been correctly estimated as the highest communication delay between the server and each of the other servers in its group or in its neighboring groups, and no message was lost, than all the commands that were supposed to be delivered before $m$ have necessarily been received already. If the same has occurred for all the servers, then all of them have received all the commands, and can deliver them in the same order of $ts$.

However, even if the optimistic order is the same for all the servers, it won't be valid if the conservative order is different from it. For that reason, we devised a way to make the conservative delivery order as close as possible to the optimistic one.

@@

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption A1, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that no more messages with a timestamp lower than $t$ would be sent to group $G_{recv}$. Let $sendersTo(G) =$ \mbox{$\{G' : G'\text{ is able to send a message to }G\}$}. When a server $S$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $S$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering is done, all the messages with timestamp up to $t$ can be conservatively delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a server has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

There are three possibilities for each message $C$, in the perspective of the coordinator $S_c$ of a given group $G$:

\begin{itemize}
  \item Command $C$ affects the state of objects in only one server group.

  In this case, when $C$ is optimistically delivered by the coordinator, it is inserted in a \textit{consPending} list for later being conservatively delivered via consensus in the local group.

  \item When $C$ affects objects in more than one group and it has originated in the same group $G$ of the coordinator $S_c$.

  In this case, when $C$ is optimistically delivered by the coordinator, it starts a Paxos consensus instance within its group, having not only its own group's servers as learners, but also those of the neighbor groups affected by $C$. The proposed value is the command $C$ itself, including $C.ts$. When the value is decided and learnt, the servers of every other affected group $G'$ adjust $barrier(G, G')$ to $C.ts$ and the coordinators of all the involved groups insert it into the respective \textit{consPending} lists.

  \item When $C$ affects objects in more than one group, but its group of origin is not $G$.

  In this case, when $C$ is optimistically delivered, nothing else is done by $S_c$, as it should be inserted into the \textit{consPending} list only when learnt via a consensus instance started at its group of origin -- stored in $C.src$.
\end{itemize}

The list \textit{consPending} is always sorted in ascending order of the timestamp of the commands in it. When the first command $C$ in this list is such that $C.ts < now - w(S_c)$ and $C.ts < barrier(G, G')$ for all $G'$ such that $G'$ is a neighbor of $G$, then $C$ is atomically broadcast (e.g. also with Paxos) to the group as the next command to be conservatively delivered.

The possibility not covered here is when a message is sent from a group to another one which is not its neighbour. However, it is realistic for this context to assume that objects should be close to each other in the virtual environment to interact and that, as such, this kind of interaction is not supported. Nevertheless, one simple way to provide this kind of support could be by each group waiting for the barriers of every other group in the system, instead of only those of its neighbors.

A more formal description of the protocol is given in Algorithm \ref{algorithm:optpaxos}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(val, learners)$, which initiates a Paxos \cite{paxosref} instance proposing the value $val$ with the local group as the group of acceptors and $learners$ as the group of learners; and also $Learn(val)$, which is called when a paxos instance achieves consensus. The $Learn(val)$ is called for all the processes which are supposed to learn a given proposed value once it is decided by a Paxos instance. For the sake of simplicity, we assume that, for each group, the values are learnt by consensus in the same order in which the consensus instances were initiated. This can be easily done with a sequence number given by the group's coordinator when initiating each consensus instance -- but note that we don't care if consensus instances initiated by different groups are finished in a non-deterministic order, but only those initiated in each single group; we want something like a FIFO ``learn order'', and not necessarily atomic.

Moreover, each server contains an \textit{optPending} list which contains the commands waiting to be \textit{OPT-Delivered} (optimistically delivered). The \mbox{\textit{consPending}} list contains the commands ready to be \textit{CONS-Delivered} (conservatively delivered), but which may be waiting for the clearance from the neighbor groups. Also, each command $cmd$ has at least three fields: $dst$, which is the list of groups which contain objects affected by it; $src$, which is the group to which the client who issued it is connected; and $ts$, the value of the wallclock of the server when it received the command from the client. Note that, by abuse of notation, we have `server $s \in C.dst$' instead of `server $s$~: $\exists$ group $G~\in~C.dst$ $\wedge$ $s~\in~G$.

\begin{algorithm}
\begin{distribalgo}[1]
\STATE \underline{This part of the algorithm is executed by every server $s$:}
\STATE
\INDENT{\textit{When $s$ receives a command $cmd$ from a client}}
  \STATE $cmd.ts \leftarrow getTime()$  
  \COMMENT{current wallclock value as the timestamp of $cmd$}
  \INDENT{\textbf{for all} $s' \in cmd.dst$ \textbf{do}}
    \STATE $send(s', cmd)$    
    \COMMENT{send optimistically $cmd$ to all involved servers}
  \ENDINDENT
  \STATE $optPending \leftarrow optPending$ $\cup$ $\{cmd\}$
\ENDINDENT
\STATE
\INDENT{\textit{When a server $s$ receives a command $cmd'$ from another server}}
  \IF {$cmd'.ts < getTime() - w(s)$}
    \STATE discard $cmd'$
    \COMMENT{late commands probably lead to out-of-order delivery}
  \ELSE
    \STATE $optPending \leftarrow optPending$ $\cup$ $\{cmd'\}$
  \ENDIF
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists c \in optPending$ : $getTime() > c.ts + w(s)$ $\wedge$\\$\nexists c' \in optPending: c'.ts < c.ts$}}
  \STATE OPT-Deliver(c)  
  \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
    \STATE $send(cli, \{c, \text{`optimistic'}\})$    
    \COMMENT{s sends $c$ to its clients, warning it is optimistic}
  \ENDINDENT
  \STATE $optPending \leftarrow optPending \setminus \{c\}$
  \IF {$s$ is a coordinator $\wedge$ $|c.dst| = 1$}
    \STATE $consPending \leftarrow consPending$ $\cup$ $\{c\}$
  \ENDIF  
\ENDINDENT
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`deliver'}\})$}
  \STATE CONS-Deliver(c)
  \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
    \STATE $send(cli, \{c, \text{`final'}\})$    
    \COMMENT{s sends $c$ to its clients, telling it is the final order}
  \ENDINDENT
  \IF {$s$ is a coordinator}
    \STATE $consPending \leftarrow consPending \setminus \{c\}$
  \ENDIF  
\ENDINDENT
\STATE
\STATE
\STATE \underline{The next part is executed only by the coordinator $s_c$ of each group $g$:}
\STATE
\INDENT{\textit{When $\exists c \in optPending$ : $getTime() > c.ts + w(s_c)$ $\wedge$ \\ $|c.dst| > 1 \wedge c.src = g \wedge \nexists c' \in optPending : c'.ts < c.ts$}}
  \STATE $Propose(\{c, \text{`multicast'}\}, c.dst)$
  \COMMENT{saving c in the acceptors before multicasting}
\ENDINDENT
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`multicast'}\})$}
  \STATE $consPending \leftarrow consPending$ $\cup$ $\{c\}$ 
  \IF {$c.src \neq g$}
    \STATE $barrier(c.src, g) \leftarrow c.ts$
  \ENDIF  
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists c \in consPending$ : \\ $c.ts < getTime() - w(s_c)$ $\wedge$ \\ $\forall g' : g'$ is a neighbor of $g$, $c.ts < barrier(g',g)$ $\wedge$ \\ $\nexists c' \in consPending : c'.ts < c.ts$ }}
  \STATE $Propose(\{c, \text{`deliver'}\}, g)$
\ENDINDENT

\caption{Delivery algorithm}
\label{algorithm:optpaxos}
\end{distribalgo}
\end{algorithm}

The problem with Algorithm \ref{algorithm:optpaxos} is that it does not guarantee liveness when a group has no message to receive from some other group and then keeps waiting for a new message to increase the barrier value and proceed with the conservative delivery. However, it does not disrupt the optimistic delivery and liveness for the conservative delivery can be easily provided by sending periodic empty commands to neighbors to which no message has been sent for a specified time threshold \textit{barrierThreshold}. Algorithm \ref{algorithm:liveness} describes this.

\begin{algorithm}
\begin{distribalgo}[1]
\STATE \underline{This algorithm is executed only by the coordinator $s_c$ of each group $g$:}
\STATE
\INDENT{\textit{When} $Learn(\{c, \text{`multicast'}\})$ $\wedge$ $c.src = g$}
  \INDENT{\textbf{for all} $g' : g' \in c.dst$ \textbf{do}}
    \STATE $lastBarrierSent(g') \leftarrow c.ts$
  \ENDINDENT
\ENDINDENT
\STATE
\INDENT{\textit{When $\exists g' : g'$ is a neighbor of $g$ $\wedge$\\$getTime() - lastBarrierSent(g') > barrierThreshold$}}
  \STATE $ec \leftarrow$ empty command
  \STATE $ec.ts \leftarrow getTime()$
  \STATE $ec.src \leftarrow g$
  \STATE $Propose(\{ec, \text{`multicast'}\}, \{g,g'\})$
  \COMMENT{saving that nothing was sent until ec.ts}
\ENDINDENT
\STATE

\caption{Achieving liveness}
\label{algorithm:liveness}
\end{distribalgo}
\end{algorithm}

@@

\subsubsection{Recovering from mistakes}

Unfortunately, even with a very good delay estimation (e.g. on an environment with a low jitter), there is absolutely no guarantee that the optimistic delivery order will match the conservative one. When it doesn't, it is considered a \emph{mistake}. Every mistake of the optimistic algorithm -- either a lost command message, or an out-of-order delivery -- will cause a rollback of the optimistic state of the objects and re-execution of some of the optimistically delivered commands.

>>>>

\newcommand{\blankline}[1]{\vspace{4 mm}}

@@

We define $sendersTo(G)$ as the set of groups which are able to send a message to $G$. Also, we define $receiversFrom(G)$ as the set of groups who are able to receive a message from $G$.

As messages are delivered according to their generation time, the delivery order is FIFO. Besides, as every two messages which are delivered in different groups should be delivered in the same order in these groups, we need total order. However, FIFO total order is not enough to guarantee one property that we need: let the send time $st(m)$ of a message $m$ be the value of the wallclock of $p$ when $m$ was sent, where $p$ is the process which sent it. We need that, if $st(m_1) < st(m_2)$, then $m_1$ should be delivered before $m_2$, if $m_1$ and $m_2$ are delivered by the same process. Although this property may be hard to observe, it is used by the optimistic delivery protocol, so that messages can be delivered in order in one communication step, in the best case. If, when each process sends a message $m$, it stamps the message with the current value $st(m)$ of its wallclock, we define:

\textbf{Global FIFO order}: if processes $p_i$ and $p_j$ send respectively $m_i$ and $m_j$ to the same destination $p_d$, then if $st(m_i) < st(m_j)$, $m_i$ is not delivered after $m_j$ in $p_d$.

Here, we consider that there are several processes. Each process $p$ belongs to a group $G = group(p)$, that is, $p \in G = group(p)$. Each group $G$ has a special leader process $c$, which is its coordinator.

@@

The basic idea of the optimistic ordering is the following: assuming that the processes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enough, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a process $p$ (either coordinator or not) receives a message $m$ from a client, it immediately applies a timestamp $ts$ to it, which consists simply of the current value of $p$'s wallclock, $now$. Therefore, $m.ts=now$. A wait window of length $w(p)$ is considered, where $w(p)$ is defined as the highest value of the estimated communication delay plus the wallclock deviation between the process $p$ and any of the other processes in its group $G$ or in any of the groups in $sendersTo(G)$.

More formally, let $\delta(p',p)$ be the maximum time for a message from $p'$ to arrive at $p$. Also, let $\epsilon(p',p)$ be the deviation between the wallclocks of $p$ and $p'$. The value of $w(p)$ is estimated as the maximum value of $(\delta(p',p) + \epsilon(p',p))$ for every $p'$ in $G = group(p)$ or in $sendersTo(G)$.

After applying the timestamp to $m$, the process $p$ immediately forwards it to all the other processes involved, including those in other groups. A process is involved with a message $m$ when it is one of its destinations, which can be inferred from $m.dst$. Then, $p$ puts $m$ in an \mbox{\textit{optPending}} list, where it stays until $now>m.ts+w(p)$, which means that $m$ has been in that list for a time longer than the defined wait window. In the meantime, other messages, sent from other processes, may have been received and also inserted in that list, always sorting by their timestamps.

If $w(p)$ has been correctly estimated, and no message was lost, than all the messages that were supposed to be delivered before $m$ have necessarily been received already. If the same has occurred for all the processes, then all of them have received all the messages, and can deliver them in the same order of $ts$.

To avoid out-of-order deliveries, when a message $m$ arrives too late, that is, if when $m$ arrives at $p$, $m.ts < now - w(p)$, the message is simply discarded by $p$, since another message $m' : m'.ts > m.ts$ may have already been delivered\footnote{We could make this in a way such that $m$ is only discarded by $p$ if, in fact, there was a delivered message $m' : m'.ts > m.ts$. If there was no such message, $m$ could still be delivered without violating the order we defined.}.

However, even if the optimistic order is the same for all the processes, it won't be valid if the conservative order is different from it. For that reason, we devised a way to make the conservative delivery order as close as possible to the optimistic one.

@@

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption A1, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that it would send no more messages with a timestamp lower than $t$ to group $G_{recv}$. We have defined that $sendersTo(G) =$ \mbox{$\{G' : G'\text{ is able to send a message to }G\}$}. When a process $p$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $p$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering (the ordering of messages originated in $G$) is done, all the messages with timestamp up to $t$ can be conservatively delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a process has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

There are three possibilities for each message $m$, in the perspective of the coordinator $c$ of a given group $G$:

\begin{itemize}
  \item The message $m$ was originated in $G$, which is the only destination of $m$:

  In this case, when $m$ is optimistically delivered by the coordinator $c$, $c$ also atomically broadcasts $m$ in $G$. When it is atomically delivered by a process $p$ of group $G$, it is not immediately delivered to the application. Instead, it is inserted in a \textit{consPending} list for later being conservatively delivered, once every group $G'$ in $sendersTo(G)$ have already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

  \item When $m$ is originated in $G$, but it has at least one group other then $G$ as a destination:

  In this case, when $m$ is optimistically delivered by the coordinator, it is also atomically broadcast, which is done by running a Paxos consensus instance within $G$, but having as learners the processes of the groups in $m.dst$. The proposed value is $\{m, barrier(G, G')\}$, where $barrier(G,G')$ is simply the value of $m.ts$. When the consensus is achieved, both the message and the barrier from $G$ are learnt by the processes of every destination group $G'$. The processes of all the involved groups also insert $m$ into the respective \textit{consPending} list.

  \item When $G$ is one of the destinations of $m$, but $m$ was originated in some other group $G'$:

  In this case, when $m$ is optimistically delivered by a process $p$ of group $G$, nothing else is done. The message $m$ should be inserted into the \mbox{\textit{consPending}} list of $p$ only when learnt via a consensus instance started at its group of origin -- stored in $m.src$.
\end{itemize}

The messages in the \textit{consPending} list are always sorted in ascending order of their timestamps. When the first message $m$, in the $consPending$ list of a process $p : group(p) = G$, is such that $m.ts < barrier(G, G')$ for all $G' \in sendersTo(G)$, then $m$ is conservatively delivered by $p$ to the application as the next message. We claim that this delivery respects the ``global FIFO total order''\footnote{Proof needed.}.

%The possibility not covered here is when a message is sent from a group to another one which is not its neighbour. However, it is realistic for this context to assume that objects should be close to each other in the virtual environment to interact and that, as such, this kind of interaction is not supported. Nevertheless, one simple way to provide this kind of support could be by each group waiting for the barriers of every other group in the system, instead of only those of its neighbors.

A more formal description of the protocol is given in Algorithm \ref{algorithm:quasi}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(val, learners)$, which initiates a Paxos \cite{paxosref} instance proposing the value $val$ with the local group as the group of acceptors and $learners$ as the group of learners; and also $Learn(val)$, which is called when a Paxos instance achieves consensus. The $Learn(val)$ is called for all the processes which are supposed to learn a given proposed value once it is decided by a Paxos instance. For the sake of simplicity, we assume that, for each group, the values are learnt by consensus in the same order in which the consensus instances were initiated\footnote{This can be easily done with a sequence number given by the group's coordinator when initiating each consensus instance -- but note that we don't care if consensus instances initiated by different groups are finished in a non-deterministic order, but only those initiated in each single group; we want something like a ``FIFO learn order'', and not necessarily atomic.}.

Moreover, each process $p$ of group $G$ keeps an \textit{optPending} list which contains the messages waiting to be \textit{OPT-Delivered} (optimistically delivered). The \mbox{\textit{consPending}} list contains the messages ready to be \textit{CONS-Delivered} (conservatively delivered), but which may be waiting for the clearance from the groups in $sendersTo(G)$. Also, each message $m$ has at least three fields: $dst$, which is the set of $m$'s destination groups; $src$, which is the group where $m$ was generated; and $ts$, the value of the wallclock of the process when $m$ was generated. Note that, by abuse of notation, we have `process $p \in m.dst$' instead of `process $p$ : $\exists$ group $G \in m.dst \wedge p \in G$'.

Something that must be noticed is that some messages might not have the source group as a destination. Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

\begin{algorithm}
\begin{distribalgo}[1]
%\blankline\
%\STATE\underline{This algorithm is executed by every process $p$, which belongs to group $G$:}
%\STATE
\blankline\
\INDENT {Initialization}
  \INDENT{\textbf{for all} $G' : G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) = -\infty$
  \ENDINDENT
\ENDINDENT
\blankline\
\INDENT{\textit{When $p$ generates a message $m$}}
  \STATE $m.ts \leftarrow getTime()$  
  \COMMENT{current wallclock value as the timestamp of $m$}
  \INDENT{\textbf{for all} $p' \in m.dst \cup G$ \textbf{do}} \label{algline:sendcoord}
    \STATE $send(p', m)$    
    \COMMENT{send optimistically $m$ to all involved processes}
  \ENDINDENT
\ENDINDENT
\blankline\
\INDENT{\textit{When a process $p$ receives a message $m'$}}
  \IF {$m'.ts < getTime() - w(p)$}
    \STATE discard $m'$
    \COMMENT{late commands probably lead to out-of-order delivery}
  \ELSE
    \STATE $optPending \leftarrow optPending$ $\cup$ $\{m'\}$
  \ENDIF
\ENDINDENT
\blankline\
\INDENT{\textit{When $\exists m \in optPending : getTime() > m.ts + w(p)$ $\wedge$\\$\nexists m' \in optPending: m'.ts < m.ts$}}
  \STATE $optPending \leftarrow optPending \setminus \{m\}$
  \IF {$p$ is a coordinator $\wedge$ $m.src = G$}
    \STATE $Propose(m, m.dst \cup G)$ \label{algline:propg}
  \ENDIF
  \IF {$G \in m.dst$}  \label{algline:notinterested}
    \STATE OPT-Deliver($m$)  
%   \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
%     \STATE $send(cli, \{c, \text{`optimistic'}\})$    
%     \COMMENT{s sends $c$ to its clients, warning it is optimistic}
%   \ENDINDENT
  \ENDIF  
\ENDINDENT
\blankline\
\INDENT{\textit{When} $Learn(m)$}
  \IF {$G \in m.dst$} \label{algline:checkcons}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \IF {$m.src \neq G$}
    \STATE $barrier(m.src, G) = m.ts$
  \ENDIF
%   \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
%     \STATE $send(cli, \{c, \text{`final'}\})$    
%     \COMMENT{s sends $c$ to its clients, telling it is the final order}
%   \ENDINDENT
%   \IF {$p$ is a coordinator}
%     \STATE $consPending \leftarrow consPending \setminus \{c\}$
%   \ENDIF  
\ENDINDENT
\blankline\
\INDENT{\textit{When $\exists m \in consPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$ $\wedge$ \\ $\nexists m' \in consPending : m'.ts < m.ts$ }}
  \STATE CONS-Deliver($m$)

\blankline\
\ENDINDENT

\caption{Delivery algorithm, executed by every process $p$ from group $G$}
\label{algorithm:quasi}
\end{distribalgo}
\end{algorithm}

\subsubsection{Addressing liveness}

The problem with Algorithm \ref{algorithm:quasi} is that it does not guarantee liveness when a group has no message to receive from some other group and then keeps waiting for a new message to increase the barrier value and proceed with the conservative delivery. However, it does not disrupt the optimistic delivery. Also, liveness for the conservative delivery can be easily provided by sending periodic empty messages from $G$ to each $G' \in receiversFrom(G)$ to which no message has been sent for a specified time threshold \textit{barrierThreshold}. Algorithm \ref{algorithm:liveness} describes this. For the coordinator of $G$ to know when a message was successfully delivered to other group via consensus, it must also learn it. For this reason, when it proposes any message for conservative delivery, it always includes $G$ as one of the group of learners (l. \ref{algline:propg} of Algorithm \ref{algorithm:quasi}). To avoid the conservative delivery of messages whose destinations do not include $G$, it checks before including it in the conservative delivery queue (l. \ref{algline:checkcons} of Algorithm \ref{algorithm:quasi}). This check also avoids delivering empty messages sent by other groups with the sole purpose of providing liveness.

\begin{algorithm}
\begin{distribalgo}[1]
\blankline\
\INDENT {Initialization}
  \INDENT{\textbf{for all} $G' : G' \in receiversFrom(G)$ \textbf{do}}
    \STATE $lastBarrierSent(G') = -\infty$
  \ENDINDENT
\ENDINDENT
\blankline\
\INDENT{\textit{When} $Learn(\{m\})$ $\wedge$ $m.src = G$}
  \INDENT{\textbf{for all} $G' : G' \in m.dst$ \textbf{do}}
    \STATE $lastBarrierSent(G') \leftarrow m.ts$
  \ENDINDENT
\ENDINDENT
\blankline\
\INDENT{\textit{When $\exists G' \in receiversFrom(G):$\\$getTime() - lastBarrierSent(G') > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow getTime()$
  \STATE $null.src \leftarrow G$
  \STATE $Propose(null, \{G,G'\})$
  \COMMENT{saving that nothing was sent until em.ts}
\ENDINDENT
\blankline\

\caption{Achieving liveness; executed by the coordinator $c$ of group $G$}
\label{algorithm:liveness}
\end{distribalgo}
\end{algorithm}

The problem with addressing liveness this way is that, in the worst case, $G'$ has decided a message $m$ and has just received the last barrier $b$ from $G$, such that $b$ is less than $m.ts$ by an infinitesimal difference. This would mean that, if $G$ has no messages to send to $G'$, $G'$ will have to wait for $barrierThreshold + 2\Delta$, assuming that $2\Delta$ is the time needed for a proposal to be learnt after it has been initiated. Because of that, the conservative delivery of $m$ will be delayed by $w(c') + 2\Delta + barrierThreshold + 2\Delta$, where $w(c')$ is the wait window of the coordinator of $G'$, and $2\Delta$ is counted twice because there are two consensus proposals -- first in $G'$ regarding $m$, and then in $G$ regarding the empty message.

There is a way to provide liveness withouth such a possibly big delay, although creating more messages and making deeper changes in the delivery algorithm. Let $blockers(m)$ be defined as the set of groups whose barrier is needed in order to some group to deliver $m$. More formally $blockers(m) = \{G_B : \exists G_{dst} \in m.dst \wedge G_B \in sendersTo(G_{dst})\}$. The idea is that, once each group $G_B$ in $blockers(m)$ have sent a barrier $b > m.ts$ to all the groups in $m.dst \cap receiversFrom(G_B)$, all possible destinations of $m$ can deliver it. This way, instead of relying on periodic messages, whenever a process $p$ in a group $G$ has a message $m$ to send -- no matter who are the destinations --, $p$ sends $m$ to every $p' \in m.dst \cup G \cup blockers(m)$. Like in Algorithm \ref{algorithm:quasi}, the message must be addressed to $G$, so that $G$'s coordinator starts a consensus instance to deliver it conservatively; and it is sent to the groups in $blockers(m)$, so that they know that there is a message which will be blocked until they send a proper barrier to unblock it.

When the coordinator $c'$ of some group $G'$ receives $m$, such that $m.src \neq G'$, $c'$ knows that there might be other groups depeding on $G's$ barrier to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equals to $m.ts$ and insert it into its optimistic queue. As soon as $null.ts > now - w(c')$, the $null$ message will be proposed by $c'$ in $G'$, having \mbox{$m.dst \cap receiversFrom(G')$} as its learners -- this way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it learns the $null$ message which was just created. The new delivery algorithm would be as described in Algorithm {\ref{algorithm:ondemand}}.

\begin{algorithm}
\begin{distribalgo}[1]
%\blankline\
%\STATE\underline{This algorithm is executed by every process $p$, which belongs to group $G$:}
%\STATE
\blankline\
\INDENT {Initialization}
  \INDENT{\textbf{for all} $G' : G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) = -\infty$
  \ENDINDENT
\ENDINDENT
\blankline\
\INDENT{\textit{When $p$ generates a message $m$}}
  \STATE $m.ts \leftarrow getTime()$  
  \COMMENT{current wallclock value as the timestamp of $m$}
  \INDENT{\textbf{for all} $p' \in m.dst \cup G \cup blockers(m)$ \textbf{do}} \label{algline:askclearance}
    \STATE $send(p', m)$
  \ENDINDENT
\ENDINDENT
\blankline\
\INDENT{\textit{When a process $p$ receives a message $m'$}}
  \IF {$G \neq m'.src$} \label{algline:askedforclearance}
    \STATE $null \leftarrow$ empty message
    \STATE $null.src \leftarrow G$
    \STATE $null.ts \leftarrow m'.ts$
    \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
    \STATE $optPending \leftarrow optPending \cup null$
  \ELSIF {$m'.ts < getTime() - w(p)$}
    \STATE discard $m'$
    \COMMENT{late commands probably lead to out-of-order delivery}
  \ELSE
    \STATE $optPending \leftarrow optPending$ $\cup$ $\{m'\}$
  \ENDIF
\ENDINDENT
\blankline\
\INDENT{\textit{When $\exists m \in optPending : getTime() > m.ts + w(p)$ $\wedge$\\$\nexists m' \in optPending: m'.ts < m.ts$}}
  \STATE $optPending \leftarrow optPending \setminus \{m\}$
  \IF {$p$ is a coordinator $\wedge$ $m.src = G$}
    \STATE $Propose(m, m.dst)$ %\label{algline:propg}
  \ENDIF
  \IF {$m \neq null$}  %\label{algline:notinterested}
    \STATE OPT-Deliver($m$)  
%   \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
%     \STATE $send(cli, \{c, \text{`optimistic'}\})$    
%     \COMMENT{s sends $c$ to its clients, warning it is optimistic}
%   \ENDINDENT
  \ENDIF  
\ENDINDENT
\blankline\
\INDENT{\textit{When} $Learn(m)$}
  \IF {$m \neq null$} %\label{algline:checkcons}
    \STATE $consPending \leftarrow consPending \cup \{m\}$
  \ENDIF
  \IF {$m.src \neq G$}
    \STATE $barrier(m.src, G) = m.ts$
  \ENDIF
%   \INDENT{\textbf{for all} $cli \in clients(s)$ \textbf{do}}
%     \STATE $send(cli, \{c, \text{`final'}\})$    
%     \COMMENT{s sends $c$ to its clients, telling it is the final order}
%   \ENDINDENT
%   \IF {$p$ is a coordinator}
%     \STATE $consPending \leftarrow consPending \setminus \{c\}$
%   \ENDIF  
\ENDINDENT
\blankline\
\INDENT{\textit{When $\exists m \in consPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$ $\wedge$ \\ $\nexists m' \in consPending : m'.ts < m.ts$ }}
  \STATE CONS-Deliver($m$)

\blankline\
\ENDINDENT

\caption{Delivery algorithm, executed by every process $p$ from group $G$}
\label{algorithm:ondemand}
\end{distribalgo}
\end{algorithm}

Now, assuming that $w(c)$ is the same for every group coordinator $c$, we have a worst-case delivery latency for $m$ of $w(c) + 2\Delta$, as the $null$ message is created and proposed in parallel with $m$. However, each sender process is sending $m$ to every destination process -- which will not be changed --, but also to every process in its group and to every process belonging to some group of $blockers(m)$. This can create an amount of messages too high to be practical. A solution for that would be the following: instead of sending $m$ to every single process in $G$ and in each group of $blockers(m)$, the sender $p$ sends $m$ only to its destinations and to the coordinators of $G$ and of the groups in $blockers(m)$. That way, the line \ref{algline:askclearance} of Algorithm \ref{algorithm:ondemand} would be:
\\
\\
\textbf{for all}\\$p' \in m.dst \cup \{c : c\text{ coordinates }G \vee (\exists G_B \in blockers(m) \wedge c\text{ coordinates }G_B)\}$\\\textbf{do}

@@

\subsection{Recovering from mistakes}

Unfortunately, even with a very good delay estimation (e.g. on an environment with a low jitter), there is absolutely no guarantee that the multicast protocol described  in section \ref{sec:quasi} will deliver the game command messages optimistically and conservatively in the same order. When it doesn't, it is considered a \emph{mistake}. Every mistake of the optimistic delivery -- either a lost command message, or an out-of-order delivery -- will cause a rollback of the optimistic state of the objects and re-execution of some of the optimistically delivered commands.
